Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2024-03-25 cis-perfect-testing.tex
fix refs
proof and format numbers

When estimating incidence from prevalence, the mean duration, not its median, is the most important feature of the duration distribution~\autocite{freemanPrevalence}.
Duration distributions tend to be right-skewed.
Therefore, estimates of the mean would be sensitive to the tail of the duration distribution.

@@

The first complication in the analysis is that infection episodes can be undetected.
Undetected infection episodes are episodes with no associated positive tests.
Therefore, there is no information directly observed about the episode; even the number that occur is latent.
They occur when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
Undetected episodes are important because they are, on average, shorter than detected episodes; therefore, neglecting them would lead to a biased estimate of the survival distribution.
To see why this is the case, consider the infection episode in figure \cref{perf-test:fig:truncation}.
Here two individuals are shown, with identical data collected for each of them (a series of negative
tests).
The first individual was never infected whereas the second individual has an undetected infection. 
The undetected infection episode begins at time 33 and has a 10-day duration.
The individual would have been detected at their next test if their duration had instead been 24 days or more.
The gap between tests in the CIS (up to 28 days, see \cref{E-intro:sec:cis}) is longer than most infections (median 16 days, see \cref{E-ATACCC:sec:results}).
Therefore, it is likely that many infections in the CIS cohort were undetected.
This is because the interval between tests in the same individual (up to 28 days, see \cref{E-intro:sec:cis}) is longer than most infections (median 16 days, see \cref{E-ATACCC:sec:results}).
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[width=\textwidth]{cis-perfect-testing/truncation}}
  \caption[Undetected episodes in CIS data]{Graphical illustration of an undetected infection episode, represented by the shaded area. \label{perf-test:fig:truncation}}
\end{figure}

@@

For this conditional likelihood to be valid, they assume this conditioning provides negligible information on the distribution of interest~\citePersonalComms{Nick Jewell}.
This assumption performs well in these settings, when the length of the intervals is fairly constant.
In particular, when the interval lengths are near-constant, $p(\text{episode $j$ detected} \mid b_j \in [l_j^{(b)}, r_j^{(b)}]) \approx p(\text{episode $j$ detected})$.
However, the interval lengths change from one to four weeks in the CIS (see \cref{E-intro:sec:cis}).

Infection episodes in an individual's one-week testing period are over-represented in the detected infections.
This is because more frequent testing will detect a higher proportion of infections.
Therefore, for the average detected episode, $p(\text{episode $j$ detected} \mid b_j \in [l_j^{(b)}, r_j^{(b)}]) > p(\text{episode $j$ detected})$.
Therefore, conditioning on $b_j \in [l_j^{(b)}, r_j^{(b)}]$ is inappropriate.

Doubly interval censored and truncated data are common when assessing the survival time of bird and insect nests~\autocite{heiseyABCs}.
\textcite{heiseyModelling}, building on long-standing methodology~\autocite{dempsterMaximum,turnbullEmpirical}, develops a likelihood-based inference framework, allowing for arbitrary truncation patterns and double interval censoring.
They apply it to a situation where the visit times (possible detection times) are common to all the individuals, simplifying the likelihood.
I adapt this framework for CIS, when each individual has their own visit schedule (see section \cref{perf-test:sec:model}).

I use Bayesian inference for two reasons.
First, the frequentist properties (\eg consistency and convergence rate) of proposed estimators are poorly understood~\autocite{sunAnalysis,dengNonparametric}.
Second, a prior can include information from previous studies, such as from \cref{E-ATACCC}.
Alternatively, the prior can favour a smoothly changing hazard, which is biologically sensible, and can overcome issues caused by the undetected episodes ~\autocite{caoBias}.

@@

In this section, I develop a statistical model accounting for double interval censoring and undetected episodes.
In \cref{perf-test:sec:data-notation}, I introduce the data and notation, defining which quantities are observed and which are latent.
Then, in \cref{perf-test:sec:episode-types}, I apply the framework of \textcite{heiseyModelling} to classify possible episode beginning and end types into \emph{admissible} episodes, which match the doubly interval censored pattern of the observations; \emph{undetected} episodes, which may have occurred but would have been undetected; and \emph{inadmissible} episodes, which the data rules out having occurred.
The number of such episodes in each class is a random variable, with a multinomial distribution, conditional on the latent total number of infections (including undetected infections).

In \cref{perf-test:sec:posterior-density}, I outline the form of the posterior density.
The following subsections then derive its various components.
Unlike in \textcite{heiseyModelling}, I must consider the probability of undetected episodes for each individual seperately because they have differing visit times.
In addition, I include the information from the individuals without a detected episode.

Unlike prior work, I marginalise the likelihood over the number of undetected episodes.
This makes inference tractable using the large sample size and the computing resources available within the SRS (see \cref{E-biology-data:sec:SRS}).
In particular, the posterior can be expressed within modern, high-performance Bayesian software such as Stan.

@@

I analyse the 4,800 episodes first detected between 10th Oct 2020 and 6th Dec 2020 inclusive, with negatives bounding the start and end time of the episode.
This period is denoted as $t = 1, \dots, T$.
A negative test before the start of the episode provides a lower bound on the episode's start time; otherwise, there is little information on its length.
Similar information comes from the negative test after the end of the episode.
A negligible number of episodes have a negative test after the end of the episode.
I refer to these episodes as \emph{detected}; in \cref{perf-test:sec:prob-undetected} I formalise the conditions for an episode to be detected.
Let $\ndet$ denote the total number of detected infections.

@@

An individual $i$ has a \emph{test schedule} $\sched_i$.
The test schedule $\sched_i$ is the set of times individual $i$ is tested at, from their last test prior to time 1 (if it exists) up until time $T$.
These are all the times an infection episode can be detected at, plus one prior test
For example, for the individual in \cref{perf-test:sec:undetected}, $\sched_i = \{ 0, 7, 14, 21, 28, 56 \}$.
Each individual has exactly one test schedule, but it is possible that $\sched_i = \sched_{i'}$ for $i \neq i'$.
I assume that the test schedules are uninformative on all quantities of interest because they are part of the study design.
Therefore, I condition on them implicitly in all probabilities that follow.

@@

To relate the possible value for $b_j$ and $e_j$, given a test schedule $\sched_{i(j)}$, to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie: are strict subsets of $T \times T$).
Figure \ref{perf-test:fig:partitionSpace} shows this graphically.
Each episode, whether detected or not, must fall into one of these three classes with a probability that is a function of the parameters of the survival function, $\vec{\vec{\theta}}$, specified in \cref{perf-test:sec:parameters-priors}.

@@

% Then $\Omega_i$ can be written as:
% \begin{align}
% \Omega_i = \{ (b, e) \ssep \tau_{\sched_i}(b) + b > e \vee b \leq \min(\sched_i) \}.
% \end{align}
Therefore, omitting the conditioning on $\vec{\theta}$ and $i(j) = i$:
\begin{align}
1 - p_{iu}
&= 1 - \prob((B_{j}, E_{j}) \in \Omega_i) \\
&= \prob(D_j \geq \tau_{\sched_{i}}(B_j)+ 1, B_j > \min \sched_{j}) \\
&= \sum_{b = \min \sched_{i} + 1}^T \prob(D_j \geq \tau_{\sched_{i}}(b) + 1 \mid B_j = b) \prob(B_j = b)\\
&= \frac{1}{T} \sum_{b = \min \sched_{i} + 1}^T S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1)
\end{align}
defining $S_{\vec{\theta}}(\infty) = 0$.

@@

I simulate a dataset of detected episodes with the same characteristics as that in the CIS by the following procedure.
\begin{enumerate}
    \item Extract the test schedules for each individual who had at least one test during the period of interest.
    \item Draw an episode start time, $b_{j}$ for each individual uniformly at random between 2nd July 2020 (100 days before the period where a detected episode would be included) and 6th December 2020 (the end of this period).
    \item Draw a duration of episode for each individual, $d_i$, based on the distribution described below. Then calculate the end of their infection episode, $e_{j} = b_{j} + d_i - 1$.
    \item Simulate the test results based on the test schedule, $b_{j}$, and $e_{j}$. Tests between $b_{j}$ and $e_{j}$ (inclusive) are positive, all other tests are negative.
    \item Discard episodes where there are no positive tests (\ie undetected episodes) and then apply the same inclusions/exclusions as in \cref{perf-test:sec:problem}.
    \item Of these remaining episodes, sample 4,800 to match the sample size of the true dataset. This is needed because in step 2 the entire cohort was infected, while in the real study only an (unknown) portion is infected.
    \item For this final set of episodes, calculate $(l_j^{(b)}, r_j^{b}, l_j^{(e)}, r_j^{(e)})$ by taking the day after the last negative prior to any positives, the first positive, the last positive, and the day before the negative following the last positive respectively.
\end{enumerate}

@@

Repeating the simulation showed negligible variation in the results (not shown) due to the large sample size (4800 detected infections).
Therefore, I show and discuss results from only a single simulation.

All analyses converged.
Rhat were below 1.01, all parameters had an effective sample size of 1000, and there were no divergent transitions.

>>>>

When estimating incidence from prevalence, the mean duration, not its median, is the most important feature of the duration distribution~\autocite{freemanPrevalence}.
Duration distributions tend to be right-skewed.
Therefore, estimates of the mean can be sensitive to the tail of the duration distribution.

@@

The first complication in the analysis is that infection episodes can be undetected.
Undetected infection episodes are episodes with no associated positive tests.
Therefore, there is no information directly observed about the episode; even the number that occur is latent.
They occur when an episode starts and ends between tests (see \cref{perf-test:fig:truncation}).
Undetected episodes are, on average, shorter than detected episodes; therefore, neglecting them would lead to a biased estimate of the survival distribution.
To see why they are shorter, consider the infection episode in \cref{perf-test:fig:truncation}.
Here two individuals are shown, with identical data collected for each of them (a series of negative
tests).
The first individual was never infected whereas the second individual has an undetected infection. 
The undetected infection episode begins at time 33 and has a 10-day duration.
The individual would have been detected at their next test if their duration had instead been 24 days or more.
The gap between tests in the CIS (up to 28 days, see \cref{E-intro:sec:cis}) is longer than most infections (median 16 days, see \cref{E-ATACCC:sec:results}).
Therefore, it is likely that many infections in the CIS cohort were undetected.
\begin{figure}
  \makebox[\textwidth][c]{\includegraphics[width=\textwidth]{cis-perfect-testing/truncation}}
  \caption[Undetected episodes in CIS data]{%
    Identical data for two individuals, with a series of negative tests.
    Top individual remains uninfected, bottom individual has an undetected infection.
    The shaded area represents when the individual is detectable.
  }
  \label{perf-test:fig:truncation}
\end{figure}

@@

For this conditional likelihood to be valid, they assume this conditioning provides negligible information on the distribution of interest~\citePersonalComms{Nick Jewell}.
This assumption performs well in these settings, when the length of the intervals is fairly constant.
A necessary condition for it to perform well, satisfied when the interval lengths are near-constant, is $p(\text{episode $j$ detected} \mid b_j \in [l_j^{(b)}, r_j^{(b)}]) \approx p(\text{episode $j$ detected})$.
However, the interval lengths change from one to four weeks in the CIS (see \cref{E-intro:sec:cis}).

Infection episodes in an individual's one-week testing period are over-represented in the detected infections.
This is because more frequent testing will detect a higher proportion of infections.
Therefore, for the average detected episode, $p(\text{episode $j$ detected} \mid b_j \in [l_j^{(b)}, r_j^{(b)}]) > p(\text{episode $j$ detected})$ and conditioning on $b_j \in [l_j^{(b)}, r_j^{(b)}]$ is inappropriate.

Doubly interval censored and truncated data are common when assessing the survival time of bird and insect nests~\autocite{heiseyABCs}.
\textcite{heiseyModelling}, building on long-standing methodology~\autocite{dempsterMaximum,turnbullEmpirical}, develops a likelihood-based inference framework, allowing for arbitrary truncation patterns and double interval censoring.
They apply it to a situation where the visit times (possible detection times) are common to all the individuals, simplifying the likelihood.
I adapt this framework for CIS, when each individual has their own visit schedule (see \cref{perf-test:sec:model}).

I use Bayesian inference for two reasons.
First, the frequentist properties (\eg consistency and convergence rate) of proposed estimators are poorly understood~\autocite{sunAnalysis,dengNonparametric}.
Second, a prior can include information from previous studies, such as \cref{E-ATACCC}.
Alternatively, the prior can favour a smoothly changing hazard, which is biologically sensible, and can overcome issues caused by the undetected episodes ~\autocite{caoBias}.

@@

In this section, I develop a computationally feasible statistical model accounting for double interval censoring and undetected episodes.
In \cref{perf-test:sec:data-notation}, I introduce the data and notation, defining which quantities are observed and which are latent.
Then, in \cref{perf-test:sec:episode-types}, I apply the framework of \textcite{heiseyModelling} which classifies possible episode beginning and end time into \emph{admissible} episodes, which match the doubly interval censored pattern of the observations; \emph{undetected} episodes, which may have occurred and been undetected; and \emph{inadmissible} episodes, which the data rules out having occurred.
The number of such episodes in each class has a multinomial distribution conditional on the latent total number of infections (including undetected infections).

In \cref{perf-test:sec:posterior-density}, I outline the form of the posterior density.
The following subsections then derive its various components.
Unlike in \textcite{heiseyModelling}, I must consider the individual's probability of undetected episodes separately because they have differing visit times.
In addition, I include the information from the individuals without a detected episode.

Unlike prior work, I marginalise the posterior density over the number of undetected episodes.
This makes inference tractable using the large sample size and the computing resources available within the SRS (see \cref{E-biology-data:sec:SRS}).
In particular, the posterior can be expressed within modern, high-performance Bayesian software such as Stan.

@@

I analyse the \numprint{4800} episodes first detected between 10 Oct 2020 and 6 Dec 2020 inclusive, with negatives bounding the start and end time of the episode.
This period is denoted as $t = 1, \dots, T$.
A negative test before the start of the episode provides a lower bound on the episode's start time; otherwise, there is little information on its length.
Similar information comes from the negative test after the end of the episode.
A negligible number of episodes have a negative test after the end of the episode.
I refer to these episodes as \emph{detected}; in \cref{perf-test:sec:prob-undetected} I formalise the conditions for an episode to be detected.
Let $\ndet$ denote the total number of detected infections.

@@

An individual $i$ has a \emph{test schedule} $\sched_i$.
The test schedule $\sched_i$ is the set of times individual $i$ is tested at, from their last test prior to time 1 (if it exists) up until time $T$.
These are all the times an infection episode can be detected at, plus one prior test.
For example, for the individual in \cref{perf-test:sec:undetected}, $\sched_i = \{ 0, 7, 14, 21, 28, 56 \}$.
Each individual has exactly one test schedule, but it is possible that $\sched_i = \sched_{i'}$ for $i \neq i'$.
I assume that the test schedules are uninformative on all quantities of interest because they are part of the study design.
Therefore, I condition on them implicitly in all probabilities that follow.

@@

To relate the possible value for $b_j$ and $e_j$, given a test schedule $\sched_{i(j)}$, to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie are strict subsets of $T \times T$).
Figure \ref{perf-test:fig:partitionSpace} shows this graphically.
Each episode, whether detected or not, must fall into one of these three classes with a probability that is a function of the parameters of the survival function, $\vec{\vec{\theta}}$, specified in \cref{perf-test:sec:parameters-priors}.

@@

% Then $\Omega_i$ can be written as:
% \begin{align}
% \Omega_i = \{ (b, e) \ssep \tau_{\sched_i}(b) + b > e \vee b \leq \min(\sched_i) \}.
% \end{align}
Therefore, omitting the conditioning on $\vec{\theta}$ and $i(j) = i$:
\begin{align}
1 - p_{iu}
&= 1 - \prob((B_{j}, E_{j}) \in \Omega_i) \\
&= \prob(D_j \geq \tau_{\sched_{i}}(B_j)+ 1, B_j > \min \sched_{j}) \\
&= \sum_{b = \min \sched_{i} + 1}^T \prob(D_j \geq \tau_{\sched_{i}}(b) + 1 \mid B_j = b) \prob(B_j = b)\\
&= \frac{1}{T} \sum_{b = \min \sched_{i} + 1}^T S_{\vec{\theta}}(\tau_{\sched_{i}}(b) + 1)
\label{perf-test:eq:piu}
\end{align}
defining $S_{\vec{\theta}}(\infty) = 0$.

@@

I simulate a dataset of detected episodes with the same characteristics as that in the CIS by the following procedure.
\begin{enumerate}
    \item Extract the test schedules for each individual who had at least one test during the period of interest.
    \item Draw an episode start time, $b_{j}$ for each individual uniformly at random between 2 July 2020 (100 days before the period where a detected episode would be included) and 6 December 2020 (the end of this period).
    \item Draw a duration of episode for each individual, $d_i$, based on the distribution described below. Then calculate the end of their infection episode, $e_{j} = b_{j} + d_i - 1$.
    \item Simulate the test results based on the test schedule, $b_{j}$, and $e_{j}$. Tests between $b_{j}$ and $e_{j}$ (inclusive) are positive, all other tests are negative.
    \item Discard episodes where there are no positive tests (\ie undetected episodes) and then apply the same inclusions/exclusions as in \cref{perf-test:sec:problem}.
    \item Of these remaining episodes, sample \numprint{4800} to match the sample size of the true dataset. This is needed because in step 2 the entire cohort was infected, while in the real study only an (unknown) portion is infected.
    \item For this final set of episodes, calculate $(l_j^{(b)}, r_j^{b}, l_j^{(e)}, r_j^{(e)})$ by taking the day after the last negative prior to any positives, the first positive, the last positive, and the day before the negative following the last positive respectively.
\end{enumerate}

@@

Repeating the simulation showed negligible variation in the results (not shown) due to the large sample size (\numprint{4800} detected infections).
Therefore, I show and discuss results from only a single simulation.

All analyses converged.
Rhat were below 1.01, all parameters had an effective sample size of \numprint{1000}, and there were no divergent transitions.

### 2024-03-25 introduction.tex
proof and format numbers

A second prevalence survey ran in the UK, known as REACT (REal-time Assessment of Community Transmission)~\autocite{rileyResurgence,rileyREACT}.
REACT was a series of cross-sectional surveys, run across 19 rounds between May 2020 and March 2022.
Each round consisted of at least 100,000 randomly sampled individuals from the UK contributing swabs for testing and questionnaire data~\autocite{elliottTwin}.

@@

\Textcite{ealesCharacterising} estimated the duration of positivity from REACT data.
Specifically, individuals who tested positive in round 8 of REACT (6th--22nd Jan 2021) were invited to take two additional swab tests.
Defining the time of the initial positive test at time 0, \textcite{ealesCharacterising} estimated the probability of remaining positive at each day after the initial positive test.
They made a strong parametric assumption that the probability decreases exponentially, with a sensitivity analysis assuming a plateau before the decay.
This assumption is necessary because they had at most 17 days of follow-up, and hence could not estimate the tail of the distribution.
Overall, they estimate a median duration of 14.0 days (95\% CI: 12.9--15.4 days) but with substantial positive correlation with each individual’s viral load at the time of the initial positive test (\ie a higher initial viral load was associated with a longer duration of positivity)
The individuals with the highest viral load individuals were estimated to have a median duration of 19.6 days (16.1--25.8 days).
The association could be due to individuals with a higher initial viral load being (on average) more recently infected at the time of the initial positive test.
The study design prohibits estimating how long before the initial positive test the infection occurred.
The same issue means the study underestimates the duration of positivity.

@@

In order to achieve these aims, I use a case study, specifically the SARS-CoV-2 epidemic in England from Mon 31st Aug 2020 until Sun 25th Jan 2021.
This case study has several features that make it useful for the aims of this thesis.
These include that the CIS was of sufficient size and incidence sufficiently high that there is enough data to perform the analyses.
Additionally, incidence begins rising over this period, which is the period when a surveillance system needs to provide a warning of upcoming issues.
The CIS began in April 2020, when incidence was falling due to the first lockdown in the UK.
Incidence remained low and fairly flat in England over summer 2020, until it began rising in mid or late August.
Finally, the period ends before the roll-out of the vaccine, which likely affects many important aspects of the analysis.
In particular, the duration of positivity and the susceptibility of the population.
Including both of these elements would require more complex modelling but include idiosyncratic features of the SARS-CoV-2 pandemic that are probably less generalisable to future diseases.

>>>>

A second prevalence survey ran in the UK, known as REACT (REal-time Assessment of Community Transmission)~\autocite{rileyResurgence,rileyREACT}.
REACT was a series of cross-sectional surveys, run across 19 rounds between May 2020 and March 2022.
Each round consisted of at least \numprint{100000} randomly sampled individuals from the UK contributing swabs for testing and questionnaire data~\autocite{elliottTwin}.

@@

\Textcite{ealesCharacterising} estimated the duration of positivity from REACT data.
Specifically, individuals who tested positive in round 8 of REACT (6th--22 Jan 2021) were invited to take two additional swab tests.
Defining the time of the initial positive test at time 0, \textcite{ealesCharacterising} estimated the probability of remaining positive at each day after the initial positive test.
They made a strong parametric assumption that the probability decreases exponentially, with a sensitivity analysis assuming a plateau before the decay.
This assumption is necessary because they had at most 17 days of follow-up, and hence could not estimate the tail of the distribution.
Overall, they estimate a median duration of 14.0 days (95\% CI: 12.9--15.4 days) but with substantial positive correlation with each individual’s viral load at the time of the initial positive test (\ie a higher initial viral load was associated with a longer duration of positivity)
The individuals with the highest viral load individuals were estimated to have a median duration of 19.6 days (16.1--25.8 days).
The association could be due to individuals with a higher initial viral load being (on average) more recently infected at the time of the initial positive test.
The study design prohibits estimating how long before the initial positive test the infection occurred.
The same issue means the study underestimates the duration of positivity.

@@

In order to achieve these aims, I use a case study, specifically the SARS-CoV-2 epidemic in England from Mon 31 Aug 2020 until Sun 25 Jan 2021.
This case study has several features that make it useful for the aims of this thesis.
These include that the CIS was of sufficient size and incidence sufficiently high that there is enough data to perform the analyses.
Additionally, incidence begins rising over this period, which is the period when a surveillance system needs to provide a warning of upcoming issues.
The CIS began in April 2020, when incidence was falling due to the first lockdown in the UK.
Incidence remained low and fairly flat in England over summer 2020, until it began rising in mid or late August.
Finally, the period ends before the roll-out of the vaccine, which likely affects many important aspects of the analysis.
In particular, the duration of positivity and the susceptibility of the population.
Including both of these elements would require more complex modelling but include idiosyncratic features of the SARS-CoV-2 pandemic that are probably less generalisable to future diseases.

### 2024-03-25 thesis.tex
proof and format numbers

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{floatpag}
\usepackage{bm}
\usepackage[algochapter]{algorithm2e}
\usepackage[nameinlink]{cleveref}
\usepackage[style=british]{csquotes}
\usepackage[T1]{fontenc}
\usepackage[unicode,psdextra]{hyperref}
\usepackage{textcomp} % provide symbols
\usepackage[disable]{todonotes}
\usepackage{xr}
\usepackage{pdflscape}
\usepackage{afterpage}
\usepackage{caption}
\usepackage{numprint}
% \usepackage{microtype}
\usepackage[UKenglish]{babel}
\usepackage{makecell}

@@

\preface{%
    In October 2019 I began a PhD project entitled ``Advancing Bayesian Statistical Methods for Real-Time Epidemic Monitoring''.
    So began a journey of reading, and an expectation of few deadlines, a lot of deep thinking, and complex statistical theory.
    Like much of the world, the events of early 2020 changed that.
    Following a brief interlude to provide rapid support to the response, the origins of this thesis emerged.

    Several aspects of my work were done to support the ongoing response.
    In particular, an earlier version of the backcalculation work was adopted by the Office for National Statistics to estimate incidence, published in their weekly online reports.
    This method was also used for the analysis of several respiratory diseases over winter 2022/3~\autocite{dietzSARSCoV2final}.

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{floatpag}
\usepackage{bm}
\usepackage[algochapter]{algorithm2e}
\usepackage[nameinlink]{cleveref}
\usepackage[style=british]{csquotes}
\usepackage[T1]{fontenc}
\usepackage[unicode,psdextra]{hyperref}
\usepackage{textcomp} % provide symbols
\usepackage[disable]{todonotes}
\usepackage{xr}
\usepackage{pdflscape}
\usepackage{afterpage}
\usepackage{caption}
\usepackage{numprint}
\npfourdigitnosep
\npdecimalsign{.}
% \usepackage{microtype}
\usepackage[UKenglish]{babel}
\usepackage{makecell}

@@

\preface{%
    In October 2019 I began a PhD project entitled ``Advancing Bayesian Statistical Methods for Real-Time Epidemic Monitoring'', aiming to develop methods in preparation for a future influenza pandemic.
    So began a journey of reading, and an expectation of few deadlines, a lot of deep thinking, and complex statistical theory.
    Like much of the world, the events of early 2020 rapidly changed that.
    Following a brief interlude to provide rapid support to the response, the origins of this thesis emerged.

    Several aspects of my work were done to support the ongoing response.
    In particular, an earlier version of the backcalculation work was adopted by the Office for National Statistics to estimate incidence, published in their weekly online reports~\autocite{cisMethodsONS}.
    This method was also used for the analysis of several respiratory diseases over winter 2022/3~\autocite{dietzSARSCoV2final}.

### 2024-03-25 transmission.tex
proof and format numbers

A slow increase in incidence is estimated to occur over the autumn 2020 period, peaking in late November in most regions (see \cref{transmission:fig:backcalc-regions}).
An exception is the North East and East of England regions (see \cref{transmission:fig:backcalc-regions}(C) and (D) respectively); these regions show no consistent increase or decrease in incidence during October.
All regions then decrease during November; this coincides with the national lockdown which came into force in England on 5th Nov 2020~\autocite{ifgTimeline}.

@@

All results are based on four Markov chains each of length 1,000,000 iterations.
The first 500,000 were discarded as burn-in, with the remaining chains thinned to every 100th iteration.
This gives 5,000 posterior samples per region.
This was further thinned to 2,000 samples when calculating posterior predictive quantities (incidence and prevalence) due to their computational cost.

@@

Finally, I generated one dataset of test results for each parameter set.
These were generated according to the likelihood, \cref{SEIR:eq:likelihood}.
The number of daily tests was set to the mean of what CIS performed in the region over this period (1630), distributed across the population in proportion to its size.

@@

All regions' MCMC chains converged, assessed using the Rhat statistic and ESS (as defined in \cref{E-MCMC:sec:convergence}).
The highest Rhat was 1.01 for $\sigma_\epsilon$ in London.
All parameters had ESS > 1000, except $\sigma_\epsilon$ and $\psi$ in London, which had ESS of 450 and 950 respectively.

>>>>

A slow increase in incidence is estimated to occur over the autumn 2020 period, peaking in late November in most regions (see \cref{transmission:fig:backcalc-regions}).
An exception is the North East and East of England regions (see \cref{transmission:fig:backcalc-regions}(C) and (D) respectively); these regions show no consistent increase or decrease in incidence during October.
All regions then decrease during November; this coincides with the national lockdown which came into force in England on 5 Nov 2020~\autocite{ifgTimeline}.

@@

All results are based on four Markov chains each of length \numprint{1000000} iterations.
The first \numprint{500000} were discarded as burn-in, with the remaining chains thinned to every 100th iteration.
This gives \numprint{5000} posterior samples per region.
This was further thinned to \numprint{2000} samples when calculating posterior predictive quantities (incidence and prevalence) due to their computational cost.

@@

Finally, I generated one dataset of test results for each parameter set.
These were generated according to the likelihood, \cref{SEIR:eq:likelihood}.
The number of daily tests was set to the mean of what CIS performed in the region over this period (\numprint{1630}), distributed across the population in proportion to its size.

@@

All regions' MCMC chains converged, assessed using the Rhat statistic and ESS (as defined in \cref{E-MCMC:sec:convergence}).
The highest Rhat was 1.01 for $\sigma_\epsilon$ in London.
All parameters had ESS > \numprint{1000}, except $\sigma_\epsilon$ and $\psi$ in London, which had ESS of 450 and 950 respectively.
